---
title: "ü6"
author: "Ben Gutzeit"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(MASS)
library(dplyr)
library(ggplot2)
library(forcats)
data(birthwt)
bwt <- birthwt %>%
mutate(
race = factor(race, levels = c(1,2,3),
labels = c("White","Black","Other")),
smoke = factor(smoke, levels = c(0,1),
labels = c("No","Yes")),
ht = factor(ht, levels = c(0,1),
labels = c("No","Yes")),
ui = factor(ui, levels = c(0,1),
labels = c("No","Yes")),
low = factor(low, levels = c(0,1),
labels = c("No","Yes"))
)
```

# 1.1 In your own words, describe what one row of the object bwt represents

-   **low** = Low birth weight
-   **age** = Mother’s age in years
-   **lwt** = Mother’s weight at last menstrual period (in pounds)
-   **race** = Mother’s race (White, Black, or Other)
-   **smoke** = Whether the mother smoked during pregnancy (Yes/No)
-   **ptl** = Number of previous premature labors
-   **ht** = History of hypertension (Yes/No)
-   **ui** = Presence of uterine irritability (Yes/No)
-   **ftv** = Number of physician visits during the first trimester
-   **bwt** = Baby’s birth weight in grams

# 1.2 Identify which variables are categorical and which are numeric.

```{r categoricalornumeric}

df <- data.frame(
  categorical = c("low","race","smoke","ht","ui"),
  numeric = c("age","lwt","ptl","ftv","bwt")
)

df
```

# 1.3 For each categorical variable, write down the meaning of its levels.

low

-   Yes = low birth weight

-   No = normal birth weight

race

-   White = white skintone

-   Black = black skintone

-   Other = other skintone

smoke

-   Yes = mother is smoker

-   No = mother is no smoker

ht

-   Yes = history of hypertension

-   No = no history of hypertension

ui

-   Yes = uterine irritability

-   No = no uterine irritability

# 2.1 Create a bar chart for and low using ggplot2.

für was und low?

# 2.2 Briefly comment on the proportion of low birth weight (low = "Yes")

```{r ue22}

lyes <- sum(bwt$low == "Yes")
lno <- sum(bwt$low == "No")
lall <- sum(bwt$low != "")

lyes <- lyes / lall
lno <- lno / lall

ldf <- data.frame(
  yes = lyes, 
  no = lno
  )

ldf

#frequency table
table(bwt$low)

```

\~ 31% of the babies are born with underweight

# 3.1 Produce histograms of bwt with three different bin widths, e.g. 100, 250, and 500.

```{r ue31}

hist(bwt$bwt, breaks = 10 , main="Age with breaks = 10")
hist(bwt$bwt, breaks = 25 , main="Age with breaks = 25")
hist(bwt$bwt, breaks = 50 , main="Age with breaks = 50")

ggplot(bwt, aes(x = bwt)) +
  geom_histogram(binwidth = 100) +
  ggtitle("Age with width = 100 and ggplot")
  

ggplot(bwt, aes(x = bwt)) +
  geom_histogram(binwidth = 250)+
  ggtitle("Age with width = 250 and ggplot")

ggplot(bwt, aes(x = bwt)) +
  geom_histogram(binwidth = 500)+
  ggtitle("Age with width = 500 and ggplot")
```

# 3.2 Describe how the choice of bin width changes the appearance of the histogram.

"breaks" changes the widths of the bars

# 3.3 Does the distribution of birth weight appear roughly symmetric, right-skewed, or leftskewed?

-   roughly symmetric

# 4.1 Plot the empirical cumulative distribution function (ECDF) of bwt using stat_ecdf().

```{r ue41}

ggplot(bwt, aes(x = bwt)) + stat_ecdf()

```
